---
title: "Visualizing Viral Integration with virolink"
format:
  html:
    toc: true
    number-sections: true
vignette: >
  %\VignetteIndexEntry{Visualizing Viral Integration with virolink}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

`virolink` organizes visualization around the questions asked of viral
integration data. The main workflow is:

1. standardize integration records
2. inspect their distribution across the host genome
3. connect host breakpoints with positions and features on the virus genome
4. move from cohort-wide patterns to individual loci

The bundled example is deliberately small. All files are loaded from the
installed package, and local chromosome sizes are used so the vignette does
not depend on network access.

## Setup

```{r}
#| label: setup
#| include: false
#| message: false
#| warning: false

library(virolink)

extdata_path <- function(file) {
  pkg_path <- system.file("extdata", file, package = "virolink")
  if (!nzchar(pkg_path)) {
    stop("Could not locate extdata file: ", file, call. = FALSE)
  }
  pkg_path
}

chrom_file <- extdata_path("chrom_sizes.txt")
chrom_color_file <- extdata_path("chrom_color.txt")
integration_file <- extdata_path("data.txt")
host_feature_file <- extdata_path("host_features.txt")

chrom_colors <- read.delim(chrom_color_file, sep = "\t", stringsAsFactors = FALSE)
host_colors <- stats::setNames(chrom_colors$color, sub("^chr", "", chrom_colors$chr))

integrations <- read_integrations(integration_file)
host_feature_table <- read.delim(host_feature_file, sep = "\t", stringsAsFactors = FALSE)
host_features_demo <- host_features(host_feature_table)
host <- host_genome(host = NULL, chrom_file = chrom_file, colors = host_colors)
virus <- c(HBV = 3215)
hbv <- hbv_features()
```

## Standardize integration records

`read_integrations()` validates the required columns and returns a reusable
integration object. The same workflow accepts an in-memory data frame through
`as_integrations()`.

```{r}
#| label: inspect-data

integrations
```

## Compare samples or cohorts

Start at the cohort level when multiple samples or groups are present. The
cohort comparison view summarizes integration burden and can stack the total by
another explicit column such as tissue, region, or any user-defined category.

```{r}
#| label: cohort-comparison-demo
#| fig-width: 7
#| fig-height: 4
#| message: false
#| warning: false

plot_cohort_comparison(
  integrations,
  group_by = "sample",
  measure = "events"
)
```

## Survey the host-genome landscape

Start with the host genome. A binned landscape reveals chromosomes and regions
with recurrent integration events without first committing to a host-virus
link layout.

```{r}
#| label: landscape-demo
#| fig-width: 10
#| fig-height: 4.8
#| message: false
#| warning: false

plot_integration_landscape(
  integrations,
  host = host,
  mode = "density",
  group_by = "sample",
  bin_size = 5e7
)
```

Use `mode = "sites"` when supporting-read depth and exact event positions are
more important than regional counts.

## Map host and virus breakpoints

The breakpoint map keeps the host genome on a continuous x-axis and places the
viral breakpoint coordinate on the y-axis. It is a compact way to ask whether
host-side recurrence and virus-side recurrence are coupled.

```{r}
#| label: breakpoint-map-demo
#| fig-width: 10
#| fig-height: 4.8
#| message: false
#| warning: false

plot_breakpoint_map(
  integrations,
  host = host,
  virus = virus,
  features = hbv,
  group_by = "sample",
  size_by = "support_reads"
)
```

## Zoom into a host locus

Once a region or breakpoint looks interesting, inspect the local context. The
locus view focuses on one host coordinate, shows nearby integration evidence,
and can overlay local annotations supplied as a small feature table.

```{r}
#| label: locus-context-demo
#| fig-width: 10
#| fig-height: 3.8
#| message: false
#| warning: false

plot_locus_context(
  integrations,
  chr = "2",
  pos = 62717949,
  window = 5000,
  annotations = host_features_demo,
  group_by = "sample",
  size_by = "support_reads"
)
```

## Examine signature and clonality evidence

Integration mechanism and clonality usually require additional experimental
context. As a first evidence view, `plot_integration_signature()` summarizes
candidate recurrent breakpoint clusters or viral-strand orientation without
turning them into automatic biological calls.

```{r}
#| label: signature-demo
#| fig-width: 8
#| fig-height: 4
#| message: false
#| warning: false

plot_integration_signature(
  integrations,
  mode = "clonality",
  group_by = "sample",
  cluster_by = "host_virus",
  cluster_window = 1000,
  measure = "support_reads",
  top_n = 8
)
```

## Connect host and virus coordinates

After surveying the host landscape, use a linked view to examine which host
breakpoints connect to which positions on the viral genome.

### Circular linked view

Track order controls stacking from the outside toward the center.

```{r}
#| label: circos-demo
#| fig-width: 10
#| fig-height: 8
#| message: false
#| warning: false

plot_integrations(
  integrations = integrations,
  host = host,
  virus = virus,
  tracks = list(
    track_ideogram(axis = "virus"),
    track_sites(
      color = "sample",
      size = "support_reads",
      label = "Sites"
    ),
    track_virus_genes(
      hbv,
      label = TRUE,
      label_min_width = 300
    ),
    track_virus_density(
      label = "Virus density"
    ),
    track_links(color = "sample")
  ),
  visual_ratio = 0.12
)
```

### Linear linked view

The linear plot gives a simple one-dimensional view of the same integration
records. It is often useful when you want a quick sanity check or a figure that
is easier to read in a report.

```{r}
#| label: linear-demo
#| fig-width: 10
#| fig-height: 4.5
#| message: false
#| warning: false

plot_linear_integrations(
  integrations = integrations,
  host = host,
  virus = virus,
  features = hbv
)
```

## Explore interactively

For HTML reports, `plot_interactive_explorer()` pairs a `ggiraph` breakpoint
map with a searchable `DT` table. Hover over points to inspect breakpoint
details, then use the table to filter the same records by sample, coordinate,
or supporting-read count.

```{r}
#| label: interactive-demo
#| message: false
#| warning: false

plot_interactive_explorer(
  integrations,
  host = host,
  virus = virus,
  features = hbv,
  group_by = "sample",
  size_by = "support_reads"
)
```

## Workflow notes

- Use `read_integrations()` for the standardized input format.
- Use `host_genome()` and `virus_genome()` when you want to make the input
  explicit and reusable.
- Use `plot_integration_landscape()` before linked views to establish the
  host-genome distribution.
- Build plots with `track_*()` objects and pass them to `plot_integrations()`.
